---
title: "Progress Memo 1"
subtitle: |
  | Final Project 
  | Data Science 1 with R (STAT 301-1)
author: "Fiona Liu"
pagetitle: "PM1 Fiona Liu"
format:
  html:
    toc: true
    toc-depth: 4
    toc-location: left
    embed-resources: true
    link-external-newwindow: true
execute:
  warning: false
from: markdown+emoji
reference-location: margin
citation-location: margin
editor_options: 
  chunk_output_type: console
---

::: {.callout-important collapse=true}
## Memo format

Like all other submissions for this course, we expect the memo to be **neat and well organized**.

A memo is a short, polished report. Meaning, it should be well written and thoughtfully laid out. Some neatness and organizational hints/tips (not an exhaustive list):

- There should be absolutely no code in your report --- don’t hide it; code should not be visible/accessible.
- Include a date indicating when report was created.
- There should be no tibbles/data frames or raw R output --- should use tables (avoid unnecessarily long output)!
- Any plots or graphics should be polished (appropriate labels and/or captions).
- All figures and tables should be properly labeled and referenced --- must make use of quarto cross-referencing.
- Should use heading and sub-headings appropriately.
- Remove prompts and instructions meant to assist students.
- Should include a link to project’s GitHub repo.
- GitHub repo should be well organized and have a few commits (no templates, commit of most recent work)


:::

::: {.callout-warning collapse=true}
GitHub (free account) cannot store large files. If you try to commit and push a large dataset you will have an ERROR! Any file over 100 MB (100,000 KB) needs to be added to the `.gitignore` file BEFORE committing.

**We may need to do that for files in your final project!**
:::


::: {.callout-tip icon=false}

## Github Repo Link

To link to your github **repo**sitory, appropriately edit the example link below. Meaning replace `https://your-github-repo-url` with your github repo url. Suggest verifying the link works before submitting.

[Fiona's Github](https://github.com/stat301-1-2025-fall/final-project-1-fionaliu0105)

:::

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(knitr)

affective_data <- read_csv("data/BRM-emot-submit.csv")
```

## Data source

The dataset for this project is a collection of affective (emotional) norms for 13,915 English words, known as lemmas.^[Warriner, A. B., Kuperman, V., & Brysbaert, M. (2013). Norms of valence, arousal, and dominance for 13,915 English lemmas. *Behavior Research Methods*, 45(4), 1191–1207.]

This data was collected via large-scale crowdsourcing, where participants rated each word on three distinct emotional dimensions:

  * **Valence:** The pleasantness of the word (from negative/unpleasant to positive/pleasant).
  * **Arousal:** The intensity of the emotion evoked (from calm/subdued to excited/energized).
  * **Dominance:** The sense of control associated with the word (from submissive/controlled to dominant/in-control).


### Why this data

I am interested in a cognitive science dataset, and this one is a good fit for an Exploratory Data Analysis (EDA). It directly relates to my interest in how humans process language and memory.

The data's structure is simple, clean, and well-documented, making it perfect for applying the data manipulation (`dplyr`) and visualization (`ggplot2`) skills from STAT 301-1. It allows for a meaningful analysis of linguistic patterns without requiring the complex machine learning or modeling techniques that are outside the scope of this course.

## Data quality & complexity check

The dataset is of high quality and low complexity.

1.  **Format & Size:** The data is available as a single, standard `.csv` file. The file size is very small (approximately 3.6 MB), which is well below the 100 MB GitHub limit noted in the course instructions . It will not require the use of `.gitignore`.
2.  **Structure:** The data is tidy, with each row representing a single word.
3.  **Key Variables:** The primary variables for analysis will be:
      * `Word`: The English word (a character string).
      * `V.Mean.Sum`: The mean valence rating (numeric).
      * `A.Mean.Sum`: The mean arousal rating (numeric).
      * `D.Mean.Sum`: The mean dominance rating (numeric).
      * Other columns, such as standard deviations for each rating, are also available and can be used to explore the consistency of the ratings.

A summary of these three numeric dimensions was generated using `dplyr::summarize()` and is shown in @tbl-data-summary. As the table indicates, there are no missing (`NA`) values in the key data columns.
```{r}
#| label: tbl-data-summary
#| tbl-cap: "Summary Statistics for Key Affective Dimensions"
#| echo: false
affective_data |> 
  select(
    Word, 
    Valence = V.Mean.Sum, 
    Arousal = A.Mean.Sum, 
    Dominance = D.Mean.Sum
  )|> 
  select(Valence, Arousal, Dominance) |> 
  pivot_longer(
    cols = everything(), 
    names_to = "Dimension", 
    values_to = "Score"
  ) |> 
  group_by(Dimension) |> 
  summarize(
    Mean = mean(Score),
    `Std. Dev.` = sd(Score),
    Min = min(Score),
    Max = max(Score),
    Missing = sum(is.na(Score))
  ) |> 
  mutate(
    Dimension = fct_relevel(Dimension, "Valence", "Arousal", "Dominance")
  ) |>
  arrange(Dimension) |> 
  kable(digits = 2)
```

```{r}
#| echo: false
#| label: fig-valence-dist
#| fig-cap: "Distribution of Mean Valence Scores (V.Mean.Sum) for 13,915 English Words"

ggplot(affective_data, aes(x = V.Mean.Sum)) +
  geom_histogram(binwidth = 0.25, fill = "steelblue", color = "white", alpha = 0.8) +
  labs(
    title = "Distribution of Word 'Valence' (Pleasantness)",
    x = "Mean Valence Rating (1=Negative, 9=Positive)",
    y = "Count of Words"
  ) +
  theme_minimal()
```


## Potential data issues

This dataset is pre-cleaned and standardized, so no significant data cleaning is anticipated. The main "challenge" will not be cleaning data but rather exploring it effectively.

The primary task will be to use functions from `dplyr` to filter and summarize the 13,915 words to answer interesting questions (e.g., "What are the 10 most positive-valence words?" or "How does arousal relate to valence?"). The analysis will focus on data exploration and visualization rather than data wrangling.





This is the perfect place for that outline. The memo instructions (from your previous prompt) state that if the data is already tidy, you're expected to provide a more "extensive and in-depth EDA report."

The best place to put your plan is in the `Potential data issues` section, as it's not really a "data issue" but a "plan of action" given the lack of issues.

I've replaced your original `Potential data issues` section with a new one called **`Potential Data Issues & Analysis Plan`**. This new section incorporates the full EDA outline we discussed, referencing the figures from the paper. I also added the `date` to the top, as your memo instructions require it.

Here is the complete, updated `.qmd` file for you to copy and paste.

## Analysis Plan

My proposed analysis plan includes:

  * **1. Univariate Analysis:** I will first explore the shape of the three main dimensions. This involves:

      * Plotting histograms for `Valence`, `Arousal`, and `Dominance` to examine their distributions.
      * Calculating summary statistics and using `dplyr::slice_max()` and `slice_min()` to find the top/bottom 10 most extreme words for each dimension.

  * **2. Bivariate Analysis:** The core of the analysis will be exploring the relationships *between* the dimensions, as shown in **Figure 3** of the paper.

      * I will plot `Valence` vs. `Arousal` (using `geom_point` and `geom_smooth`) to find the "U-shaped" relationship, which shows that both very positive and very negative words are highly arousing.
      * I will also plot `Valence` vs. `Dominance` to show their linear relationship.

  * **3. Group Difference Analysis:** I will explore the demographic data by comparing the `V.Mean.M` (Male) and `V.Mean.F` (Female) columns, as inspired by **Figures 10-12**.

      * I plan to create scatter plots comparing male and female ratings directly, using `geom_abline` to show the line of equality.
      * I will also `filter()` for specific "taboo" or "weapon" words (as mentioned in the paper) to see if I can replicate the finding that males rate these words as more positive and in-control.

  * **4. Variability Analysis:** Finally, I will analyze rating *agreement* by plotting the mean scores (e.g., `V.Mean.Sum`) against their standard deviations (`V.SD.Sum`), replicating **Figure 2** to see if neutral words are indeed the most "ambiguous" (have the highest standard deviation).

This plan will allow me to heavily use `dplyr` for filtering and summarizing and `ggplot2` for creating a wide variety of polished visualizations, demonstrating mastery of the course's core tools.

```
```